Numerical simulation of the Cordilleran ice sheet through the last glacial cycle

### 2014-06-16 cordillera-cycle.tex
Add missing references in model setup.

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model \aref. The model inputs basal topography, sea level, geothermal
heat flux and climate forcing, and computes the evolution of ice extent
and thickness in time, the thermal and dynamic state of the ice sheet, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{Bueler.Brown.2009,Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.Blatter.2009,Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Artemieva.Mooney.2001,Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{Aschwanden.etal.2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments
\citep{Martin.etal.2011,Aschwanden.etal.2013}. The till effective pressure is
related to the modelled amount of water in the till, using a formula derived
from laboratory experiments with till extracted from an Antarctic ice stream
\aref. Basal topography is derived from the ETOPO1 combined topography
and bathymetry dataset with a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

@@

The positive degree-day integral \citep{Calov.Greve.2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability associated
with the
seasonal cycle itself \citet{Seguinot.Rogozhina.2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study \aref.

@@

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project (GRIP), the North Greenland Ice Core Project
(NGRIP), the European Project for Ice Coring in Antarctica (EPICA), the Lake
Vostok ice core, and Ocean Drilling Program (ODP) sites 1012 and 1020, both
located off the Californian shore \aref. Palaeo-temperatures from the GRIP and
NGRIP records were calculated using a quadratic equation \aref,
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided by their authors \aref. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~kyr after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Source\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C & \aref \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C & \aref \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C & \aref \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C & \aref \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}

>>>>

The simulations presented here were run using Parallel Ice Sheet Model (PISM,
development version~11b0a7f), an open-source, finite-difference, shallow ice
sheet model \citep{PISM-authors.2014}. The model inputs basal topography,
sea level, geothermal
heat flux and climate forcing, and computes the evolution of ice extent
and thickness in time, the thermal and dynamic state of the ice sheet, and
the associated lithospheric response. Our modelling domain encompasses the
entire area covered by the Cordilleran ice sheet at the last glacial maximum
(Fig.~\ref{fig:locmap}).

@@

PISM is a~shallow model, which implies that the balance of stresses is
approximated based on their predominant components.
The Shallow Shelf Approximation (SSA) is used as a ``sliding law'' for the
Shallow Ice Approximation (SIA) \citep{Bueler.Brown.2009,Winkelmann.etal.2011}.
Ice softness depends on temperature and water content through an enthalpy
formulation \citep{Aschwanden.Blatter.2009,Aschwanden.etal.2012}. Surface air
temperature from the atmospheric forcing provides the upper boundary condition
to the ice enthalpy model. Temperature is computed subglacially to a~depth of
3\,km, where it is conditioned by a~geothermal heat flux of
70\,\unit{mW\,m^{-2}}. Although this uniform value does not
account for the high spatial geothermal variability in the region, it is on
average representative of available heat flow measurements
\citep{Blackwell.Richards.2004}.

A~pseudo-plastic sliding law \citep{Aschwanden.etal.2013} relates the
bed-parallel shear stress and the sliding velocity. The yield stress is
modelled using the Mohr--Coulomb criterion. The till friction angle $\phi$
varies from 15 to 45{\degree}. It is taken as a~function of modern bed
elevation, with lowest values occurring at low elevation, thus accounting
for a~weakening of till associated with the presence of marine sediments.
The till effective pressure is related to the modelled amount of water in the
till, using a formula derived from laboratory experiments with till extracted
from an Antarctic ice stream \citep{Tulaczyk.etal.2000}. Basal topography is
derived from the ETOPO1 combined topography and bathymetry dataset with
a~resolution of 1\,arc-min \citep{Amante.Eakins.2009}.
Sea level is lowered as a function of time according to the SPECMAP
reconstruction, and basal topography responds to ice load
following a bedrock deformation model that includes point-wise isostasy,
elastic lithosphere flexure and viscous mantle deformation in a~semi-infinite
half-space \citep{Lingle.Clark.1985,Bueler.etal.2007}.

@@

The positive degree-day integral \citep{Calov.Greve.2005} is numerically
approximated using week-long sub-intervals. It accounts for temperature
variability assuming a~normal distribution along a~central (input) value. The
temperature standard deviation is part of the forcing climatology. It was
computed from daily temperature values from the North American Regional
Reanalysis (NARR) \citep{Mesinger.etal.2006}, after excluding variability
associated with the seasonal cycle itself \citep{Seguinot.Rogozhina.2014}. The
ablation model incorporates degree-day factors of
3.04\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for snow and
4.59\,\unit{mm\,{\degree}C^{-1}\,day^{-1}} for ice, as derived from
mass-balance measurements on contemporary glaciers from the Coast Mountains and
Rocky Mountains in British Columbia \citep{Shea.etal.2009}.

@@

The present-day climatology $\{T_{m0}, P_{m0}\}$ was computed from
near-surface air temperature and precipitation rate fields of the NARR over
the period 1979--2000 (Fig.~\ref{fig:atm}). Modern climate of the North
American Cordillera is characterised by strong geographic variations of
seasonality, timing of the precipitation peak and daily temperature variability.
Our choice of data from the NARR is motivated by the need for an accurate,
high-resolution precipitation forcing, as outlined in a previous sensitivity
study \citep{Seguinot.etal.2013}.

@@

Temperature offset time-series ${\Delta}T_{TS}$ are derived from proxy records from
the Greenland Ice Core Project \citep[GRIP,][]{Dansgaard.etal.1993}, the
North Greenland Ice Core Project \citep[NGRIP,][]{Andersen.etal.2004},
the European Project for Ice Coring in Antarctica \citep[EPICA,][]
{Jouzel.etal.2007}, the Lake Vostok ice core \citep{Petit.etal.1999}, and Ocean
Drilling Program (ODP) sites 1012 and 1020, both located off the Californian
shore \citep{Herbert.2001}. Palaeo-temperatures from the GRIP and NGRIP records
were calculated using a quadratic equation \citep{Johnsen.etal.1995},
\begin{equation}
    {\Delta}T_{TS}(t) = -11.88[\chem{\delta^{18}O(t)}-\chem{\delta^{18}O}(0)]
                        -0.1925[\chem{\delta^{18}O(t)}^2-\chem{\delta^{18}O}(0)^2],
\end{equation}
while the temperature reconstructions from the Antarctic and oceanic cores were
provided as such. All records were scaled linearly in
order to simulate realistic and comparable ice extents at the last
glacial maximum (Table~\ref{tab:records}, Fig.~\ref{fig:timeseries}).

@@

% tab:records
\begin{table*}[t]
  \caption{Palaeo-temperature proxy records and scaling parameters used to
           prepare temperature offset time-series used to force the ice sheet
           model through the last 120\,kyr. $T_{[32;22]}$ refers to the
           mean temperature anomaly during the period -32 to~-22~kyr after
           scaling.}
  \label{tab:records}
  {\begin{tabular}{lcccc}
    \tophline
    Record & Proxy & Scaling factor & $T_{[32;22]}$ & Original reference\\
    \middlehline
    GRIP     & \chem{\delta^{18}O} & ?\% & -5.8{\degree}C
        & \citet{Dansgaard.etal.1993} \\
    NGRIP    & \chem{\delta^{18}O} & ?\% & -6.1{\degree}C
        & \citet{Andersen.etal.2004} \\
    EPICA    & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
        & \citet{Jouzel.etal.2007} \\
    Vostok   & \chem{\delta^{18}O} & ?\% & -5.6{\degree}C
        & \citet{Petit.etal.1999} \\
    ODP~1012 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
        & \citet{Herbert.2001} \\
    ODP~1020 & \chem{U^{K'}_{37}}  & ?\% & -5.8{\degree}C
        & \citet{Herbert.2001} \\
    \bottomhline
  \end{tabular}}
  \belowtable{}
\end{table*}
